A Transactional Memory System for Parallel Python

### 2014-07-30 talk/dls2014/paper/paper.tex
removing a lot of footnotes and other cleanups

\emph{Threading} employs operating system (OS) threads to provide
concurrency. It is, however, limited by the GIL and thus does not
provide parallelism\footnote{At this point we should mention that it is indeed
possible to run external functions written in C instead of Python in
parallel. Our work focuses on Python itself and ignores this aspect as
it requires writing in a different language.}.  Moreover, the GIL can
only ensure correctness of the interpreter itself: applications are
required to coordinate concurrent accesses to data structures using
conventional methods -- locks being the most common way.

@@

When we store into our objects the references to other objects, we
need to be careful: addresses that are valid in one segment are not
valid in another, because they would break isolation by possibly pointing to
the copy of the object in another segment.  Consequently, we would
need to adapt these addresses to the segment they live in, which would
prevent completely any page sharing because then the binary contents
differ between segments.  Instead, we store pointers as
\emph{Segment Offsets ($SO$)}, i.e.\ offsets from the start of the
segment.  An $SO$ can then be interpreted in any segment by adding to
it the start address of that segment.\footnote{An alternative model
would be to simulate threads using several independent processes in
which all segments are at the same address.  This might work well for
demonstration purposes, but it is not really suitable for implementing
an existing programming language: programs assume that most of the
state (file descriptors, external libraries, ...) is shared.} The
result is called a \emph{Linear Address (LA)}. This is illustrated in
Figure~\ref{fig:Segment-Addressing}.

The x86 CPUs provide a feature which is called \emph{memory
segmentation}.  On the modern x86-64, this is enough to perform the
addition described above very efficiently.  We use the segment
register $\%gs$\footnote{The other segment register $\%fs$ is
typically used by the threading library to provide access to
thread-local data.  One point of view is that we are using $\%gs$ for
a similar purpose: thread-local data -- but a lot of it.}, which is
available to applications, to point to the current thread's segment
start address.  The translation from an $SO$ to an LA is done for us by
using the ``$\%gs\colon$'' prefix before any CPU instruction that
reads or writes memory using an $SO$ address.  This is very efficient
-- we can do it on every object access -- and some compilers support it
natively (e.g.\ clang).

@@

In summary, translating a $\%gs:SO$ to a physical address is a
two-step process: First the memory segmentation feature of the CPU
constructs a linear address. Then, this LA gets mapped by the MMU to
some place in the physical memory. This makes the $SO$ a valid reference
to an object in all segments, automatically resolving either to a
shared or a private version.

@@

What we described so far has no measurable performance overhead by
itself, but it is missing conflict detection to be considered a
full-fledged STM system.  This requires adding \emph{barriers} to the
program to register the objects in the read or write set before
reading or writing to them\footnote{Barriers added by an automatic, conservative, and
local program transformation~\cite{felber07}}.  Furthermore, we still
need to describe the commit protocol.

@@

\item [{Read~Barrier:}] A major contribution to our performance
  is the simplicity of the read barrier: a large program, like an
  interpreter for a complex language, contains a number of reads that
  typically vastly exceeds the number of writes.  The described
  approach allows the read barrier to be a single, unconditional
  write into segment-local memory. We only need to \emph{mark} an object
  as read by the current transaction. The mark is stored in a
  segment-local array indexed by a number derived from the $SO$
  of the object.

@@

  Here, we eagerly detect write-write conflicts by allowing only one
  transaction modifying an object at a time using write
  locks.\footnote{Eager write-write detection is not inherent to our
    approach; we may lift this restriction in the future.}  Part of making
  the object write-ready is also the privatisation of all pages the
  object belongs to, as described above.

@@

In the following sections, whenever we use $SO$, we go through the
address translation to get to the actual contents of an object. This
is also signified by the type \lstinline!object_t!.  This type is
special as it causes the
compiler\footnote{We use Clang 3.5 with patches to deal with bugs in
its ``address-space 256'' feature. Patches are available from authors
until inclusion into the official clang.} to make all accesses through
it relative to the $\%gs$ register.  With exceptions, nearly all
accesses to objects managed by the TM system use this type so
that the CPU will translate the reference to the right version of the
object.
\medskip   % why, Latex, why??

@@

Another issue is the support of
irreversible operations, like calls to external functions or general
input~/ output. If we are allowed to, we commit the transaction
before and start a new one after these operations\footnote{The language
specification needs to specify if this is allowed.}. Within an atomic
block, however, the transaction is unbreakable, and we need to turn
the transaction inevitable so that it
cannot abort and strong isolation is guaranteed.

@@

\begin{description}
\item [{NULL~page:}] This page is unmapped and will produce a
  segmentation violation when accessed. We use this to detect
  erroneous dereferencing of \lstinline!NULL! references.  All
  $\%gs{::}SO$ translated to linear addresses will point to NULL pages
  if $SO$ is set to \lstinline!NULL!.
\item [{Segment-local~data:}] Some area private to the segment that
  contains segment-local information for bookkeeping.
\item [{Read~markers:}] These are private pages that store information about
  which objects were read in the current transaction running in this
  segment.
\item [{Nursery:}] This private area contains all the freshly allocated
  objects (\emph{young objects}) of the current transaction. The GC
  uses bump-pointer allocation in this area to allocate objects in the
  first generation.
\item [{Old~object~space:}] These pages are the ones that are really
  shared between segments. They mostly contain old objects but also
  some young ones that were too big to be allocated in the nursery.
\end{description}

Note, since the above configuration is currently specified at compile time,
all these areas are at offsets inside the segments known to the
compiler. This makes some checks very efficient, e.g.\ checking
if an object resides in the nursery only requires comparing its $SO$
to the static boundaries of the nursery. It is possible that we want
some parameters to be configurable at startup or even during the
runtime. In that case we can still use a compile-time specified
maximum so that these checks are still efficient. E.g.\ limiting the
maximum amount of memory available to the application statically to a few
terabytes is fine because it corresponds to virtual memory,
the real physical memory is assigned on-demand by the operating
system.

@@

\subsection{Garbage Collection}

@@

This area can be seen as a continuous, segment-local array of bytes
that is indexed with an object's reference ($SO$) divided by 16. So
that each object has its own read marker, all objects have a size of
at least 16 bytes. Otherwise there could be false conflicts when
reading from two adjacent objects that share a single marker.
Instead of just setting the byte to \lstinline!true!  if the
corresponding object was read, we set it to a \lstinline!read_version!
belonging to the transaction, which will be incremented on each
commit.  Thereby, we can avoid resetting the bytes to
\lstinline!false!  on commit and only need to do this every 255
transactions. The whole code for the barrier shown in
Listing~\ref{lst:rb} is easily optimisable for compilers as well as
perfectly predictable for CPUs\footnote{Additional benefit: The read
barrier is not constrained to execute before the actual read -- both
the compiler and the CPU are free to reorder or group them.}.

@@

The \textbf{fast path} of the write barrier is very simple. We only
need to check for the flag \lstinline!WRITE_BARRIER! in the object's
header through its $SO$ and call the slow path if it is set. This flag
is set either if the object is old and comes from an earlier
transaction, or if there was a minor collection (in this case, the flag
is added again on all objects, including new overflow objects). The flag
is never set on young, freshly allocated objects. This fast path
covers all cases for the GC and for the STM system (see
Listing~\ref{lst:w_fast}).

@@

For the \emph{STM part}, we first perform a read barrier on the
object. We then try to acquire its write lock. \lstinline!write_locks!
is a simple, \emph{global} array of bytes that is indexed with the
$SO$ of the object divided by 16. Note, ``global'' here really means
it is a single array with data for all segments, there is no address
translation going on to access its elements contrary to e.g.\ the
read markers array.  If we already own the lock, we are done.
If someone else owns the lock, we perform  write-write contention
management that aborts either us or the current owner of the
object.  If the lock is not owned so far, and if we succeed in
acquiring it using an atomic
\lstinline!cmp_and_swap!, we need to add the object to the write set
(a simple list called \lstinline!modified_old_objects!).  We then
privatise the page or pages where it resides (copy-on-write).

In all cases, we remove the \lstinline!WRITE_BARRIER!  flag from the
object before we return. Thus, we never trigger the slow path again
before we do the next minor collection or we start the next
transaction.\footnote{We always do a minor collection during a commit.}

@@

Committing a transaction needs a bit more work. First, we synchronise
all threads so that the committing one is the only one running and all
the others are waiting in safe points.\footnote{This design choice is
important to support our very cheap barriers. It is a trade-off that
works best if the number of concurrent threads is small.
See section~\ref{subsub:sync} for how we mitigate the costs.}
We then go through the write
set (\lstinline!modified_old_objects!)  and check the corresponding
\lstinline!read_markers!  in other threads~/ segments. If we detect a
read-write conflict, we perform contention management to either abort us or
the other transaction, or to simply wait a bit (see Section~\ref{subsub:contentionmanagement}).

@@

We also ran these benchmarks on Jython\footnote{version
2.7b1~\cite{webjython}}, an implementation of Python on top of the
Java Virtual Machine (JVM).  Instead of a GIL, this interpreter uses
fine-grained locking for synchronisation. Even though it can use the
JVM's JIT compiler, its performance in these benchmarks is behind
PyPy's by a factor of $5-10\times$. Because of that it does not
make sense to include it in this evaluation. In the future, we would
like to do an extensive comparison between the STM approach and
fine-grained locking. It is out of the scope of this paper to do
this thoroughly.

>>>>

\emph{Threading} employs operating system (OS) threads to provide
concurrency. It is, however, limited by the GIL and thus does not
provide real parallelism.
The only code that is not necessarily restricted by the GIL in CPython
is code written in e.g.\ C instead of Python. External functions can
choose to run without the GIL and thus, in parallel. We ignore this
aspect here as it requires writing in a different language than Python
and is therefore not relevant to our work.

The GIL can also
only ensure correctness of the interpreter itself: applications using
threads are still
required to coordinate concurrent accesses to data structures using
conventional methods -- locks being the most common way.

@@

When we store into our objects the references to other objects, we
need to be careful: addresses that are valid in one segment are not
valid in another, because they would break isolation by possibly pointing to
the copy of the object in another segment.  Consequently, we would
need to adapt these addresses to the segment they live in, which would
prevent completely any page sharing because then the binary contents
differ between segments.  Instead, we store pointers as
\emph{Segment Offsets (SO)}, i.e.\ offsets from the start of the
segment.  An SO can then be interpreted in any segment by adding to
it the start address of that segment. The
result is called a \emph{Linear Address (LA)}. This is illustrated in
Figure~\ref{fig:Segment-Addressing}.

The x86 CPUs provide a feature which is called \emph{memory
  segmentation}.  On these CPU this is enough to perform the addition
described above very efficiently. On the modern x86-64, there are two
segment registers left: $\%fs$ and $\%gs$.  On Linux for example,
$\%fs$ is already used for addressing thread-local data
efficiently. Thus, we use the remaining segment register $\%gs$, which is
still available to applications, to point to the current thread's segment
start address.  The translation from an SO to an LA is done for us by
using the ``$\%gs\colon$'' prefix before any CPU instruction that
reads or writes memory using an SO address.  This is very efficient
-- we can do it on every object access -- and some compilers support it
natively (e.g.\ clang).

@@

In summary, translating a $\%gs:SO$ to a physical address is a
two-step process: First the memory segmentation feature of the CPU
constructs a linear address. Then, this LA gets mapped by the MMU to
some place in the physical memory. This makes the SO a valid reference
to an object in all segments, automatically resolving either to a
shared or a private version.

@@

What we described so far has no measurable performance overhead by
itself, but it is missing conflict detection to be considered a
full-fledged STM system.  This requires adding \emph{barriers} to the
program to register the objects in the read or write set before
reading or writing to them. We do this with an automatic, conservative, and
local program transformation (cf.~\cite{felber07}).  Furthermore, we still
need to describe the commit protocol.

@@

\item [{Read~Barrier:}] A major contribution to our performance
  is the simplicity of the read barrier: a large program, like an
  interpreter for a complex language, contains a number of reads that
  typically vastly exceeds the number of writes.  The described
  approach allows the read barrier to be a single, unconditional
  write into segment-local memory. We only need to \emph{mark} an object
  as read by the current transaction. The mark is stored in a
  segment-local array indexed by a number derived from the SO
  of the object.

@@

  Here, we eagerly detect write-write conflicts by allowing only one
  transaction modifying an object at a time using write
  locks\footnote{Eager write-write conflict detection is not inherent
    to our approach; we may lift this restriction in the future if it
    proves to be useful.}. To make the object write-ready, this is
  also the place to privatise all pages the object belongs to, as
  described above.

@@

In the following sections, whenever we use SO, we go through the
address translation to get to the actual contents of an object. This
is also signified by the type \lstinline!object_t!.  This type is
special as it causes the
compiler\footnote{We use Clang 3.5 with patches to deal with bugs in
its ``address-space 256'' feature. Patches are available from authors
until inclusion into the official clang.} to make all accesses through
it relative to the $\%gs$ register.  With exceptions, nearly all
accesses to objects managed by the TM system use this type so
that the CPU will translate the reference to the right version of the
object.
\medskip   % why, Latex, why??

@@

Another issue is the support of
irreversible operations, like calls to external functions or general
input~/ output. If Python allows it, we commit the transaction before
and start a new one after these operations (the same way as the GIL is
released around thread-safe external functions). Within an atomic
block, however, the transaction is unbreakable, and we need to turn
the transaction inevitable so that it cannot abort and strong
isolation is guaranteed at all times.

@@

\begin{description}
\item [{NULL~page:}] This page is unmapped and will produce a
  segmentation violation when accessed. We use this to detect
  erroneous dereferencing of \lstinline!NULL! references.  All
  $\%gs{::}SO$ translated to linear addresses will point to NULL pages
  if SO is set to \lstinline!NULL!.
\item [{Segment-local~data:}] Some area private to the segment that
  contains segment-local information for bookkeeping.
\item [{Read~markers:}] These are private pages that store information about
  which objects were read in the current transaction running in this
  segment.
\item [{Nursery:}] This private area contains all the freshly allocated
  objects (\emph{young objects}) of the current transaction. The GC
  uses bump-pointer allocation in this area to allocate objects of the
  first generation. To make sure that only objects of the current
  transaction reside in this space, a collection of the first
  generation happens at the end of every transaction (see
  section~\ref{sub:gc}).
\item [{Old~object~space:}] These pages are the ones that are really
  shared between segments. They mostly contain old objects but also
  some young ones that were too big to be allocated in the nursery.
\end{description}

Note, since the above configuration is currently specified at compile time,
all these areas are at offsets inside the segments known to the
compiler. This makes some checks very efficient, e.g.\ checking
if an object resides in the nursery only requires comparing its SO
to the static boundaries of the nursery. It is possible that we want
some parameters to be configurable at startup or even during the
runtime. In that case we can still use a compile-time specified
maximum so that these checks are still efficient. E.g.\ limiting the
maximum amount of memory available to the application statically to a few
terabytes is fine because it corresponds to virtual memory;
the real physical memory is assigned on-demand by the operating
system.

@@

\subsection{Garbage Collection\label{sub:gc}}

@@

This area can be seen as a continuous, segment-local array of bytes
that is indexed with an object's reference (SO) divided by 16. So
that each object has its own read marker, all objects have a size of
at least 16 bytes. Otherwise there could be false conflicts when
reading from two adjacent objects that share a single marker.
Instead of just setting the byte to \lstinline!true!  if the
corresponding object was read, we set it to a \lstinline!read_version!
belonging to the transaction, which will be incremented on each
commit.  Thereby, we can avoid resetting the bytes to
\lstinline!false!  on commit and only need to do this every 255
transactions. The whole code for the barrier shown in
Listing~\ref{lst:rb} is easily optimisable for compilers as well as
perfectly predictable for CPUs. Additionally, the read barrier is not
constrained to execute before the actual read -- both the compiler and
the CPU are free to reorder or group them for efficiency.

@@

The \textbf{fast path} of the write barrier is very simple. We only
need to check for the flag \lstinline!WRITE_BARRIER! in the object's
header through its SO and call the slow path if it is set. This flag
is set either if the object is old and comes from an earlier
transaction, or if there was a minor collection (in this case, the flag
is added again on all objects, including new overflow objects). The flag
is never set on young, freshly allocated objects. This fast path
covers all cases for the GC and for the STM system (see
Listing~\ref{lst:w_fast}).

@@

For the \emph{STM part}, we first perform a read barrier on the
object. We then try to acquire its write lock. \lstinline!write_locks!
is a simple, \emph{global} array of bytes that is indexed with the
SO of the object divided by 16. Note, ``global'' here really means
it is a single array with data for all segments, there is no address
translation going on to access its elements contrary to e.g.\ the
read markers array.  If we already own the lock, we are done.
If someone else owns the lock, we perform  write-write contention
management that aborts either us or the current owner of the
object.  If the lock is not owned so far, and if we succeed in
acquiring it using an atomic
\lstinline!cmp_and_swap!, we need to add the object to the write set
(a simple list called \lstinline!modified_old_objects!).  We then
privatise the page or pages where it resides (copy-on-write).

In all cases, we remove the \lstinline!WRITE_BARRIER!  flag from the
object before we return. Thus, we never trigger the slow path again
before we do the next minor collection or we start the next
transaction (remember: we always perform a minor collection as part of
each commit anyway).

@@

Committing a transaction needs a bit more work. First, we synchronise
all threads so that the committing one is the only one running and all
the others are waiting in safe points.
We then go through the write
set (\lstinline!modified_old_objects!)  and check the corresponding
\lstinline!read_markers!  in other threads~/ segments. If we detect a
read-write conflict, we perform contention management to either abort us or
the other transaction, or to simply wait a bit (see Section~\ref{subsub:contentionmanagement}).

Note that synchronising all threads is an important design choice we
made to support our very cheap read barriers. Since they now consist
only of a single unconditional write to some memory location, all
threads need to be stopped while we detect conflicts. This is a
trade-off that works best if the number of concurrent threads is
small.

@@

We also ran these benchmarks on Jython (v2.7b1 from~\cite{webjython}),
an implementation of Python on top of the
Java Virtual Machine (JVM).  Instead of a GIL, this interpreter uses
fine-grained locking for synchronisation. Even though it can use the
JVM's JIT compiler, its performance in these benchmarks is behind
PyPy's by a factor of $5-10\times$. Because of that it does not
make sense to include it in this evaluation. In the future, we would
like to do an extensive comparison between the STM approach and
fine-grained locking. It is out of the scope of this paper to do
this thoroughly.
